---
title: "COVID19 Report"
author: "Pier"
date: "2023-04-30"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


###Covid19 Report

##Importing Data

```{r import}
library(tidyverse)
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"

file_names<- c("time_series_covid19_confirmed_global.csv",
"time_series_covid19_deaths_global.csv",
"time_series_covid19_confirmed_US.csv",  
 "time_series_covid19_deaths_US.csv"
)

urls<- str_c(url_in,file_names)
urls 
#enter

global_cases <- read_csv(urls[1])
global_deaths <- read_csv(urls[2])
US_cases <- read_csv(urls[3])
US_deaths <- read_csv(urls[4])

#enter
global_cases

```

## Data Manipulation

In this section I cleaned up the data. I deleted columns that we didn't need and also changed some incorrect data types. I filtered out the days with 0 cases because they are not important to the analysis.

I also merged the data sets so that cases and deaths could be in all in one table.

```{r manipulation}
#pivot longermake each column into a row expect province,country,lat,long
#column headin into dates and values will go to cases
#select everything but lat long
global_cases <- global_cases %>%
pivot_longer(cols=-c('Province/State',
'Country/Region',Lat,Long),
names_to = "date",
values_to= "cases")%>%
select(-c(Lat,Long))

library(lubridate)

global_deaths <- global_deaths %>%
pivot_longer(cols=-c('Province/State',
'Country/Region',Lat,Long),
names_to = "date",
values_to= "deaths")%>%
select(-c(Lat,Long))


#combine cases and deaths

global <- global_cases %>%
  full_join(global_deaths)%>%
  rename(Country_Region = `Country/Region`,
         Province_State = `Province/State`) %>%
  mutate(date= mdy(date))
#gilter out dates with 0 cases 

global <- global %>% filter(cases>0)
summary(global)

global %>% filter(cases>28000000)
#enter

US_cases %>%
  pivot_longer(cols = -(UID:Combined_Key),
               names_to = "date",
               values_to = "cases")
#enter
#format the date, remove long/lat, and somehting with admin
US_cases <- US_cases%>%
  pivot_longer(cols = -(UID:Combined_Key),
               names_to = "date",
               values_to = "cases")%>%
  select(Admin2:cases) %>%
  mutate(date=mdy(date))%>%
  select(-c(Lat, Long_))


US_deaths <- US_deaths %>%
  pivot_longer(cols = -(UID:Population),
               names_to = "date",
               values_to = "deaths")%>%
  select(Admin2:deaths) %>%
  mutate(date=mdy(date))%>%
  select(-c(Lat, Long_))

#join datasets
US<- US_cases %>% 
  full_join(US_deaths)
#now i have the cases, population and deaths by date

global<- global %>%
  unite("combined_Key",
        c(Province_State, Country_Region),
        sep = ", ",
        na.rm = TRUE,
        remove = FALSE)

#add population
uid_lookup_url<- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
uid <- read_csv(uid_lookup_url) %>%
  select(-c(Lat,Long_, Combined_Key, code3, iso2,iso3,Admin2))

#join it by province state,
#select everything but UID and FIPS
#select all those fields 

global <- global%>%
  left_join(uid, by= c("Province_State", "Country_Region")) %>%
  select(-c(UID,FIPS))%>%
  select(Province_State, Country_Region, date,
         cases,deaths, Population,
         combined_Key)

global


```


```{r tidy}
#us by state grouped by ragion date
#summarize => summ of casesand deaths
#mutate add columns, death per mil and then select
#wanted features



US_by_state <- US %>% 
  group_by(Province_State, Country_Region, date)%>%
  summarize(cases=sum(cases), deaths =sum(deaths),
            Population = sum(Population) ) %>%
  mutate(deaths_per_mill = deaths*1000000/Population)%>%
  select(Province_State, Country_Region, date,
         cases, deaths, deaths_per_mill, Population)%>%
  ungroup()



#create us totat, for each date and each day the numbers of ases and deaths per million 


US_totals <- US_by_state %>% 
  group_by(Country_Region, date)%>%
  summarize(cases=sum(cases), deaths =sum(deaths),
            Population = sum(Population) ) %>%
  mutate(deaths_per_mill = deaths*1000000/Population)%>%
  select(Country_Region, date,
         cases, deaths, deaths_per_mill, Population)%>%
  ungroup()



```

##Visual 1
Visualization of total number of cases and number of Deaths in the US from COVID19

```{r visualone, echo=TRUE}
#see the end of dataset 
# take cases over 0
# x be date y num
#add line and point to graph
#scale y variable in log scale

US_totals %>%
  filter(cases>0)%>%
  ggplot(aes(x=date, y= cases))+
  geom_line(aes(color="cases"))+
  geom_point(aes(color= "cases"))+
  geom_line(aes(y=deaths, color="deaths"))+
  geom_point(aes(y=deaths, color="deaths"))+
  scale_y_log10()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90))+
  labs(title="COVID19 in USA", y=NULL)
```

##Visualization 2 

COVID19 cases and deaths in the state of New York

```{r visualtwo, echo=TRUE}
#filter by state 
state<-"New York"
US_by_state %>%
  filter(Province_State==state)%>%
  ggplot(aes(x=date, y= cases))+
  geom_line(aes(color="cases"))+
  geom_point(aes(color= "cases"))+
  geom_line(aes(y=deaths, color="deaths"))+
  geom_point(aes(y=deaths, color="deaths"))+
  scale_y_log10()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90))+
  labs(title=str_c("COVID19 in ",state), y=NULL)

```

##Info about data

As of 2023-03-09, there has been 1123836 COVID related deaths


```{r dataanalysis}
max(US_totals$date)
max(US_totals$deaths)


```

##Data Analysis

Data transfrom: add new variable to see new cases and new deaths
```{r dataanalysistwo}

US_by_state <- US_by_state %>%
  mutate(new_cases= cases-lag(cases),
         new_deaths= deaths-lag(deaths))

US_totals <- US_totals%>%
  mutate(new_cases= cases-lag(cases),
         new_deaths= deaths-lag(deaths))

```

##Visualization 3 
This visualization shows the new cases and new deaths from COVID

```{r visualizationthree, echo=TRUE}

US_totals %>%
  ggplot(aes(x=date, y= new_cases))+
  geom_line(aes(color="new_cases"))+
  geom_point(aes(color= "new_cases"))+
  geom_line(aes(y=new_deaths, color="new_deaths"))+
  geom_point(aes(y=new_deaths, color="new_deaths"))+
  scale_y_log10()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90))+
  labs(title="COVID19 in USA", y=NULL)
```

##Data Analysis

This Analysis show the top 10 state with the highest and lowest death rate per thousand people 

```{r dataanalysisthree}
#group by state and choose max death and cases 
#compute death/casses per thousand 

US_state_totals<- US_by_state%>%
  group_by(Province_State)%>%
  summarise(deaths=max(deaths), cases=max(cases),
            population=max(Population),
            cases_per_thou=1000*cases/population,
            deaths_per_thou=1000*deaths/population)%>%
  filter(cases>0, population>0)



#slice_min tell me ten state with smalles death per thousands

US_state_totals %>%
  slice_min(deaths_per_thou, n=10) %>%
select(deaths_per_thou, cases_per_thou, everything())

US_state_totals %>%
  slice_max(deaths_per_thou, n=10) %>%
  select(deaths_per_thou, cases_per_thou, everything())
```

##Linear Model

This model shows how cases are are predicting the deaths for COVID

```{r linearmodel, echo=TRUE}
#creating a linear model-one variable predicts anothe rone
#deaths a function of cases

mod<- lm(deaths_per_thou ~ cases_per_thou, data= US_state_totals)
summary(mod)
US_state_totals %>% mutate(pred= predict(mod))


US_tot_w_pred <- US_state_totals %>% mutate(pred=predict(mod))
US_tot_w_pred
```

##Visualization 4

This is a plot that shows the predicted cases from the linear model in red and also shows the real number of cases in blue 


```{r visualization, echo=TRUE}
US_tot_w_pred%>% ggplot()+
  geom_point(aes(x=cases_per_thou, y=deaths_per_thou), color="blue")+
  geom_point(aes(x=cases_per_thou, y=pred), color="red")
```
Additional questions:
1. What caused the cases with large residual? 
2. what other variables could be effecting the model?



## Global Anlysis
This sections will look at the global data for COVID19
First it will show the the global cases and global deaths from COVDId
Then It show the new cases and deaths from covid
The it will show a linear regresion between cases and deaths.



```{r tidy2}

global <- global %>% 
  group_by(Province_State, Country_Region, date)%>%
  summarize(cases=sum(cases), deaths =sum(deaths),
            Population = sum(Population) ) %>%
  mutate(deaths_per_mill = deaths*1000000/Population)%>%
  select(Province_State, Country_Region, date,
         cases, deaths, deaths_per_mill, Population)%>%
  ungroup()




```
##Visualization 5

```{r visual1, echo=TRUE}

global %>%
  filter(cases>0)%>%
  ggplot(aes(x=date, y= cases))+
  geom_line(aes(color="cases"))+
  geom_point(aes(color= "cases"))+
  geom_line(aes(y=deaths, color="deaths"))+
  geom_point(aes(y=deaths, color="deaths"))+
  scale_y_log10()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90))+
  labs(title="COVID19 Global Data", y=NULL)
```
Data transform: add new variable to see new cases and new deaths
```{r dataanalysis2}

global <- global %>%
  mutate(new_cases= cases-lag(cases),
         new_deaths= deaths-lag(deaths))

```

## Visualization 6
```{r visualizationthree, echo=TRUE}

global %>%
  ggplot(aes(x=date, y= new_cases))+
  geom_line(aes(color="new_cases"))+
  geom_point(aes(color= "new_cases"))+
  geom_line(aes(y=new_deaths, color="new_deaths"))+
  geom_point(aes(y=new_deaths, color="new_deaths"))+
  scale_y_log10()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90))+
  labs(title="COVID19 Globaly ", y=NULL)
```

Model 2



```{r visualization, echo=TRUE}
global%>% ggplot()+
  geom_point(aes(x=cases_per_thou, y=deaths_per_thou), color="blue")+
  geom_point(aes(x=cases_per_thou, y=pred), color="red")
```




##Bias disclosure 

Are the deaths or cases always reported the same?
Do I trust the data?
Why does the global linear regresion seem to show hgiher margins of errors. 


##Conclusion

In conclusion, this report presented an analysis of COVID-19 data in the United States. The report showcased the process of importing and cleaning the data, as well as visualizations of the cases and deaths related to COVID-19. The analysis of the data showed that, while the overall number of cases and deaths appeared to have flattened out, the number of new cases and deaths has remained consistent. Additionally, a linear model was created to predict the number of COVID-19 deaths based on the number of cases. While the model was generally accurate, there were some points with large residuals that require further investigation.

Overall, this report provides valuable insights into the trends of COVID-19 in the United States, highlighting the need for continued monitoring and research to understand the impact of the pandemic on public health. 

